\documentclass[11pt]{article}
\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{caption}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\captionsetup{font=small,labelfont=bf}

\title{\textbf{TurboQuant-Pro: Keep the Geometry}\\[4pt]
\large Task-Geometry-Aware Compression for Embeddings and KV Caches,\\
with Certificates Instead of Promises}
\author{Andrew H. Bond\\ San Jos\'e State University\\
\texttt{andrew.bond@sjsu.edu} \quad ORCID 0009-0003-2599-6158}
\date{Working draft --- v0.1}

\begin{document}
\maketitle

\begin{abstract}
Vector compression is usually evaluated by reconstruction fidelity and shipped
on a promise. This paper describes TurboQuant-Pro, an open-source compression
system for embedding retrieval and LLM KV caches, organized around a single
principle learned the hard way and then proved: \emph{decompose scale from
direction, and keep whichever part carries the downstream task's geometry ---
then certify it, don't promise it}. The system contributes, on the practice
side: a PCA-Matryoshka + scalar-quantization retrieval pipeline with
compressed-domain search (27$\times$ storage at high recall); the discovery
that per-vector polar quantization of attention \emph{keys} passes
reconstruction benchmarks while destroying generation (perplexity
$\sim$10$^4$ at cosine 0.995) and the per-channel architecture that fixes it;
asymmetric NormalFloat (asym-NF4), one calibration-free codebook that is
near-fp16 on every architecture tested where symmetric NF4 silently collapses
on high-ratio GQA models; the measurement that the key DC offset absorbing
that zero-point is RoPE-frequency-structured (Spearman 0.91--0.98 against
rotary wavelength; 96--99\% of offset mass in channels whose wavelength
exceeds the window), yielding a \emph{deterministic, zero-calibration
zero-point} computed from weights and config alone that matches calibrated
quality on perplexity and beats it on LongBench; and a fused decode kernel
family that computes attention directly on compressed codes --- including,
new here, per-channel keys with dense-sparse fp16 outliers fused as sparse
score deltas (exact to 8e-8, 2--6$\times$ over decompress-then-attend); and a
decision-level equivalence metric with a noise floor answering the
\emph{illusion of equivalency}, which de-confounded at matched bits reverses a
published projection-sensitivity claim --- value/output weights are $2.3$--$6\times$
more behaviorally sensitive than query/key, the keys finding's mirror image
under weight rather than activation quantization; a compiler pass that
\emph{infers} the observer from the model graph (structural rules plus
\texttt{torch.fx}) rather than requiring it declared; and the (A2) sensitivity
boundaries for the two operators beyond attention --- MoE routing, carried by the
selection margin not the logit magnitude, and state-space decay, whose slow
channels quantized in a log-time-constant basis cut real-Mamba perplexity from
$10^{10}$ to near-baseline. On
the guarantees side, imported from a companion theory paper: distribution-free
\emph{rank certificates} (Kendall $\tau \geq 1-2\hat\mu(\kappa)$ for any
corpus, with vacuity as a principled ``rerank required'' signal), and a
consumer-metric probe that reproduces the keys catastrophe at calibration
time. Two further empirical laws --- a density-quotient dissociation
(quotient density only when it is nuisance for the task truth; confirmed
out-of-sample on public data) and a spectral boundary for bit-tapering
(effective rank vs.\ bit-head width) --- complete a methodology in which every
production incident becomes an installed instrument, every claim carries an
evidence level, and negative results are published next to the wins they
bound.
\end{abstract}

\section{Introduction: the reconstruction-metric fallacy}\label{sec:intro}
The central failure this system is built around is easy to state: a quantizer
was near-perfect by every reconstruction metric --- 0.095 mean relative error,
cosine similarity 0.995 --- and raised a 7B model's perplexity from 12.2 to
10{,}643. The tensor was the attention \emph{key} cache; the quantizer
preserved each key's norm and quantized its direction; and the consumer,
$\mathrm{softmax}(QK^{\top})$, reads per-channel scale structure that
per-vector normalization deletes. Reconstruction fidelity and task fidelity
are not merely loosely coupled --- on keys they were \emph{anti-correlated}
across our quantizer family (\S\ref{sec:keys}).

The constructive lesson generalizes. Compression is the choice of a quotient:
some coordinates are kept, some are discarded or coarsened. The correct
quotient is determined not by the data's statistics but by the
\emph{downstream metric} --- what we call the task geometry. For cosine
retrieval over isotropic embeddings, direction is the geometry and per-vector
scale is disposable (store it, cheaply, exactly). For attention keys,
per-channel scale \emph{is} the geometry. For graph geodesics --- the setting
of the companion theory paper \cite{kta} --- the angular coordinate carries
everything and the radius is provably degree noise. One principle covers all
three: \textbf{keep the geometry; sometimes that is the angle, sometimes
angle plus norm, sometimes per-channel scale}. The companion paper makes the
boundary precise (its condition (A2): a scale-discarding quotient is safe
exactly when the consumer's metric is carried by the tangential part of the
displacement). The consumer-relative synthesis \cite{obstheory} sharpens this
into a coding theorem and, crucially, \emph{demonstrates} the mechanism: a
compressor that \emph{omits} a direction the consumer reads incurs an
irreducible downstream \emph{distortion floor} that no bit budget removes ---
now shown on a trained Llama attention layer, where the omitted read direction
costs a softmax divergence $800$--$5700\times$ (median $0.10$ nats) that of a
rate-matched consumer-aware coder, sealed and confirmed out-of-sample. The keys
catastrophe is that floor. This paper is the practice half --- the systems, the
discoveries the principle organizes, and the instruments that keep the next
violation from shipping.

\paragraph{Contributions.}
\begin{enumerate}
\item \textbf{A two-track compression system} (\S\ref{sec:system}):
  embedding/vector-DB compression with compressed-domain (ADC) search, and
  KV-cache compression with an asymmetric-by-quantizer key/value
  architecture; 528 tests, evidence-laddered claims, reproducible harnesses.
\item \textbf{The keys finding} (\S\ref{sec:keys}): reconstruction metrics
  cannot gate key quantization; per-channel scale is the consumer geometry.
\item \textbf{asym-NF4 and the GQA collapse mechanism} (\S\ref{sec:nf4}):
  one calibration-free codebook robust across MHA and high-GQA
  architectures, with the failure it fixes traced end-to-end.
\item \textbf{The RoPE-frequency structure of key offsets and the
  deterministic zero-point} (\S\ref{sec:rope}): the offset lives in
  config-identifiable slow rotary channels; a zero-point computed from
  weights + config alone ties calibrated quality on perplexity and beats it
  on LongBench-qasper (43.36 vs.\ 42.35; fp16 43.77).
\item \textbf{Behavior is the gate} (\S\ref{sec:behavioral}): a decision-level
  equivalence metric with a noise floor (answering the ``illusion of
  equivalency''), and a de-confounded experiment that \emph{reverses} a
  published projection-sensitivity ranking --- at matched bits value/output
  weights are functionally $2.3$--$6\times$ more sensitive than query/key
  across three models, the keys finding's mirror image under the opposite
  operator.
\item \textbf{Inferring the observer} (\S\ref{sec:infer}): a compiler pass
  (structural rules $+$ \texttt{torch.fx}) that maps each tensor to its operator
  regime and (A2) discipline from the model alone --- the declared consumer
  becomes an inferred one, so bit allocation on a novel architecture needs no
  human in the loop.
\item \textbf{Operator sensitivity beyond attention} (\S\ref{sec:hybrid}): the
  measured (A2) boundary for MoE routing (selection is carried by the margin, not
  the magnitude) and state-space decay (slow channels fragile; a
  log-time-constant basis cuts real-Mamba perplexity from $10^{10}$ to
  near-baseline).
\item \textbf{Certificates instead of promises} (\S\ref{sec:cert}):
  distribution-free rank-agreement floors from measured distortion and
  corpus concentration; a calibration-time consumer-metric probe; a
  streaming drift guardrail --- the theory paper's results shipped as
  product features.
\item \textbf{Two empirical laws with their boundaries} (\S\ref{sec:laws}):
  the density-quotient dissociation (confirmed out-of-sample) and the
  spectral effective-rank boundary for bit-tapering (a synthetic win that
  honestly reversed on public data).
\item \textbf{Compute-on-codes decode} (\S\ref{sec:fused}): a fused
  attention kernel family over compressed KV, culminating in per-channel
  keys with dense-sparse outliers fused as warp-cooperative sparse score
  deltas --- no reconstruction, no scatter, no dense-loop branching.
\item \textbf{A methodology} (\S\ref{sec:method}): evidence ladder,
  pre-registration, incident-to-instrument.
\end{enumerate}

\section{The system}\label{sec:system}
\textbf{Track 1 (embeddings).} A PCA-Matryoshka rotation (PCA recovers the
ordered eigenfunctions Matryoshka training targets \cite{varici2025}) makes
non-Matryoshka models truncatable without retraining; a random-rotation
scalar quantizer (TurboQuant \cite{turboquantpaper}) codes the unit
direction; the norm is stored beside the codes. Retrieval runs
\emph{compressed-domain}: an ADC index scores queries directly against codes
(AVX2 kernel), with exact reranking at a fixed oversample as the standard
protocol. Headline operating point: 27$\times$ storage at high recall@10,
beating RaBitQ on recall and tying OPQ at 4--20$\times$ lower index-build
cost, under one CI-tested harness on public data with provided ground truth.
The honest scope is part of the claim: PCA truncation wins on
concentrated-spectrum encoders and loses to PQ/OPQ on compact descriptors;
at matched bytes the scalar quantizer alone still wins or ties.

\textbf{Track 2 (KV caches).} A two-tier streaming cache (fp16 hot window,
compressed cold pages) that is \emph{asymmetric by quantizer}: values through
the polar flow (near-lossless), keys through \texttt{PerChannelKV}
(\S\ref{sec:keys}--\ref{sec:rope}), with dense-sparse fp16 outliers and an
attention sink. Model-aware configuration (\texttt{AutoConfig}) selects bits,
RoPE-aware boosting, and --- since the zero-point work --- the key zero-point
mode per family. A fused CUDA decode family computes attention on the codes
(\S\ref{sec:fused}).

\section{The keys finding}\label{sec:keys}
Quantizing post-RoPE keys with the per-vector polar flow (rotation, L2
normalization, Lloyd-Max codebook) yields the best reconstruction error of
our key-quantizer family and the worst generation quality by three orders of
magnitude (Qwen2.5-1.5B, WikiText-2: ppl 10{,}643 vs.\ fp16 12.24), while a
per-channel uniform quantizer with 1.5$\times$ the reconstruction error is
near-fp16. The mechanism is geometric: attention logits are bilinear in the
keys, so what must survive is per-channel scale structure across the token
axis, precisely the coordinate per-vector normalization quotients away. Two
system-level consequences: (i) \texttt{PerChannelKV} --- per-channel
asymmetric scales over tokens, optional non-uniform levels, dense-sparse
outliers --- is the shipped key path; (ii) no reconstruction metric gates any
key change in this project; generation (perplexity, then task scores) is the
gate. The finding is also a boundary result for the companion theory: keys
are the counterexample to ``always keep the angle,'' cited there as the
scope of the polar move \cite{kta}.

\section{asym-NF4 and the GQA collapse}\label{sec:nf4}
The calibration-free NF4 codebook, a common default, \emph{silently
collapses} on high-ratio GQA models: Qwen2.5-7B (7:1) drops
LongBench-qasper from 43.8 (fp16) to 4.7 and WikiText-2 perplexity from 7.46
to 74.7, while the same recipe is near-lossless on Llama-2 and Mistral ---
invisible if one benchmarks only the Llama family. The causal chain, each
link measured: KV keys carry a large per-channel DC offset, so NF4's
zero-centred abs-max grid wastes half its codes on the empty side
(representation); NF4's key error is \emph{equal} on Qwen and Llama, so the
differentiator is error \emph{tolerance} --- a 7:1 GQA model routes each
KV-head's error into seven query heads (amplification). The fix is one
per-channel zero-point: \textbf{asym-NF4} centres the grid before applying
the NF levels --- best-or-tied on every model tested at no extra bit cost
(Qwen qasper recovered to 41.9), calibration-free, and a dead heat with
KVQuant's Fisher-calibrated NUQ on LongBench. Details and the cross-model
matrix are in the satellite paper \cite{kvtmlr}.

\section{The offset is RoPE geometry; the zero-point is free}\label{sec:rope}
Where does the DC offset live? Measuring post-RoPE keys across Qwen2.5
1.5B/7B/14B and Mistral-7B: the per-channel offset magnitude correlates with
the channel's rotary wavelength at pooled Spearman 0.91--0.98, with
96--99\% of the offset mass in the channels whose wavelength exceeds the
context window (67\% of channels at $\theta{=}10^6$, 52\% at $10^4$). The
mechanism: a rotary channel slower than the window is position-near-invariant
within it, so any pre-RoPE per-channel mean survives rotation as DC, while
fast channels average it away. This closes the \S\ref{sec:nf4} chain
(structural DC $\rightarrow$ codebook waste $\rightarrow$ GQA amplification)
and makes the zero-point computable \emph{without data}: pushing the model's
own \texttt{k\_proj} bias through the position-averaged rotation gives a
zero-point from weights + config alone. Validated twice: WikiText-2
perplexity 12.533/9.179 (Qwen2.5-1.5B/7B) vs.\ 12.534/9.155 calibrated; and
LongBench-qasper \textbf{43.36} vs.\ 42.35 calibrated in the same harness
(fp16 43.77, symmetric NF4 4.69) --- the zero-calibration zero-point
\emph{beats} dense calibration on the collapse-sensitive task. A second lean
mode stores calibrated means only on the config-identified slow channels
(one-third less metadata, 42.66 qasper). Both ship in \texttt{PerChannelKV}
(\texttt{zero\_point="bias"/"sparse"}), plumbed through the streaming cache
with per-block absolute positions, and ``bias'' is the Qwen-family default in
\texttt{AutoConfig}. Scope: the bias route requires key-projection biases
(the Qwen family); bias-free models keep the calibrated or sparse modes.

\section{Behavior is the gate: the illusion of equivalency and the flipped projection}\label{sec:behavioral}
One rung above the reconstruction-metric fallacy sits a task-metric fallacy: a
quantized model can match the full-precision model's \emph{accuracy} while
getting a different \emph{set} of items right --- aggregate scores preserved,
individual decisions churned (the ``illusion of equivalency'' \cite{illusion}).
The remedy is the same one level up: gate on the consumer's decisions, with a
control. We ship a decision-level instrument, \texttt{behavioral\_agreement}: a
symmetric flip rate (McNemar regressions \emph{and} recoveries, whose sum is
the churn that accuracy nets to zero), a prediction-level agreement (do the two
models return the same answer, right or wrong), and a \emph{noise floor} --- the
churn between two near-lossless requantizations --- so drift is reported as
excess over floor with a $z$-score. This repairs two defects in the Correctness
Agreement of \cite{illusion}: it is bounded above by $\min(\mathrm{acc})$ and
blind to base-wrong$\rightarrow$quant-right recoveries, and it has no control
for the free flip rate of near-threshold items. On Qwen2.5-1.5B (next-token on
WikiText-2, all weights per-output-channel fake-quantized) a 4-bit run whose
accuracy falls $\sim$10 points hides that \textbf{41\% of token predictions
changed} (agreement 0.591; churn 16.6\% $\gg$ the 9.5\% accuracy delta),
$\sim$28$\sigma$ over a 0.014 floor. On the source paper's own Mistral-7B a
4-bit run with a 4.3-point accuracy drop hides 23\% churn (33$\times$ floor),
while 8-bit Correctness Agreement equals base accuracy --- so the 14-point
``lossless-Q8'' gap that paper reports (Llama-3.2-3B) is near-threshold item
brittleness, which the floor isolates rather than attributes to quantization.

The instrument then settles a mechanism claim. \cite{illusion} reports the
query/key projections as \emph{more} quantization-sensitive than value/output
and advises compressing V/O harder --- but measures sensitivity as
weight-distribution drift under \texttt{llama.cpp} K-quant, whose schemes give V
and O \emph{more} bits than Q/K (\texttt{Q2\_K}: Q,K at 2-bit, V at 4, O at 3):
``Q/K drift more'' is entangled with ``Q/K got fewer bits.'' De-confounded ---
the \emph{same} per-output-channel $b$-bit quantizer on all four projections,
drift read both in weight space and functionally (the movement of the output
distribution, $\mathrm{KL}(p_{\text{base}}\Vert p_{\text{quant}})$, from
perturbing one projection in every layer) --- the ranking inverts.
Weight-space drift is \emph{equal} across Q/K/V/O (relative-Frobenius ratio
$\approx 1$ at every bit-width), while functionally \textbf{V/O move the output
$2.3$--$6\times$ more than Q/K} across three architectures, including the source
paper's own Mistral-7B ($3.4\times$ at 4-bit; per-projection top-1 flip places
V and O above Q and K). The advice is backwards for weight PTQ: at matched bits
V/O are the projections whose error the model tolerates \emph{least}; Q/K only
draw level at destructive 2-bit, where the model is already broken.

The payoff is the paper's principle sharpened. This finding and the keys
finding (\S\ref{sec:keys}) look contradictory --- there keys are the fragile
side, here keys are the robust side --- and they reconcile through one
architectural fact read under two operators. $\mathrm{softmax}(QK^\top)$
mediates Q and K, so a \emph{shared} weight perturbation to $W^K$ is partly
absorbed by attention's relative comparison, while $W^V,W^O$ write the residual
stream almost linearly and pass their error on; but an \emph{activation}/KV
perturbation on the score path corrupts each cached key's per-channel scale
directly, which the softmax cannot absorb (\S\ref{sec:keys}). Same geometry,
opposite fragile coordinate, selected by the operator: keys under activation
quantization, V/O under weight quantization. Neither ordering is visible in
weight-distribution or reconstruction statistics --- skewness, kurtosis,
cosine, the source paper's own tools --- only in a consumer-metric readout.
That is the (A2) discipline of \S\ref{sec:cert} applied to weight PTQ, and one
more incident-to-instrument: the metric ships (\texttt{behavioral\_agreement},
scipy-free, unit-tested; L5), the finding is a reproducible experiment
(\texttt{experiments/matched\_bit\_projection\_sensitivity.py}; L4), and the
scope is part of the claim --- fake-quantization, a clean per-output-channel
absmax control (not any vendor kernel), three models on a WikiText-2 next-token
sample, relative and directional. Llama-3.2-3B (gated), the downstream
multiple-choice tasks, and the authors' own \texttt{llama.cpp} pipeline forced
to equal bit-width are the open confirmations.

\section{Inferring the observer}\label{sec:infer}
Every result so far names its observer by hand: a human knew keys feed
$\mathrm{softmax}(QK^\top)$ and V/O write the residual, and passed that to the
(A2) probe (\S\ref{sec:cert}) as a \emph{declared} consumer. That is the last
manual step, and \texttt{operator\_trace} removes it. Given only a model, it
maps every parameter tensor to the operator its output flows into --- one of
\{score path, linear residual, gated selection, state decay, norm\} --- and from
the regime to the (A2) discipline, with no declaration. Two front-ends combine.
A \emph{structural} pass reads module type and name (a \texttt{LayerNorm} is a
norm regardless of name; a router \texttt{gate} is selection, but SwiGLU's
\texttt{gate\_proj} is not); it is always available but blind to renamed
operators. A best-effort \texttt{torch.fx} pass symbolically traces the graph,
finds the sink operators --- softmax, top-$k$, cumulative scan --- and
backtraces to the \texttt{Linear} layers feeding them, tagging the
score/gate/state tensors \emph{even when the names are obfuscated}: on a module
whose query/key projections are named \texttt{left}/\texttt{right}, the graph
pass alone recovers them as the score path. Where fx has positive evidence it
overrides the structural guess; otherwise the structural baseline stands, and an
untraceable model degrades gracefully rather than failing.

The discipline is a function of the regime \emph{and the quantization target},
because the sensitive coordinate flips between the two --- the reconciliation of
\S\ref{sec:keys} and \S\ref{sec:behavioral} made mechanical. For the score path,
the projection's \emph{weights} are the robust, symmetric side (softmax absorbs
a shared perturbation) while its \emph{cached keys} are the fragile
per-channel-plus-zero-point side; for the residual path, the weights are the
sensitive side (V/O, \S\ref{sec:behavioral}) and the cached values are cheap
polar. An unclassified tensor defaults to the conservative per-channel
discipline --- never discard scale on a coordinate whose observer is unknown.
This is the point at which ``keep the geometry the observer sees'' stops needing
a human to name the observer.

\section{Beyond attention: gates and state decay}\label{sec:hybrid}
Hybrid and state-space models add two operators whose sensitive coordinate is
neither per-channel DC (keys) nor symmetric residual (V/O).
\texttt{operator\_sensitivity} supplies each one's (A2) analysis, and each
boundary was measured before it was believed.

\textbf{Gates: selection is carried by the margin, not the magnitude.} A top-$k$
router reads only the \emph{order} of its logits $\ell = W_g x$, so it is
invariant to a common-mode shift --- a common-mode quantization error is free,
and only the \emph{differential} component (the part orthogonal to the all-ones
direction) can flip a route. That is the tangential/(A2) split restated for
selection, and \texttt{differential\_fraction} is its statistic, the routing
analog of the tangential fraction. A token survives iff the perturbation stays
below its \emph{margin} (the gap between the $k$-th and $(k{+}1)$-th logit), so
the fragile tokens are the low-margin ones and the sensitive quantity is the
margin distribution, not the logit magnitude. On a synthetic $16$-expert router,
top-1 flips concentrate on low-margin tokens by $1.5\times$, $6.3\times$, and
$88\times$ at 2-, 3-, and 4-bit gate quantization respectively --- almost every
surviving flip at usable precision is a low-margin token.
\textbf{On real Mixtral-8x7B-Instruct-v0.1} (top-2 routing, WikiText-2,
$131{,}072$ routing decisions), the top-2 boundary margins are substantial
(median logit gap $0.485$, tenth percentile $0.076$)---unlike a high-$k$ router
such as OLMoE (top-8 of 64), whose near-zero boundary margins (median $0.0015$)
saturate under any coarse quantization. Per-tensor gate-weight quantization flips
the top-2 expert set for $16.6\%$ of tokens at 4-bit and $34.0\%$ at 3-bit, and
the damage is margin-concentrated exactly as predicted: below-median-margin
tokens flip $10.7\times$ more often than above-median at 4-bit, and $2.7\times$
at 3-bit. A controlled differential-logit perturbation isolates the mechanism
further---at a quarter of the median margin the low-margin flip rate exceeds the
high-margin rate by over $2000\times$, collapsing toward $1\times$ only as the
perturbation grows past the margins. The absolute flip rate tracks the harshness
of the gate quantizer, but the \emph{margin-concentration} is the invariant, and
it is far stronger on the substantial-margin top-2 router than the saturated
top-8 one.\footnote{Measured by \texttt{benchmarks/validate\_mixtral\_routing.py}
and \texttt{validate\_olmoe\_routing.py}; raw data in the corresponding
\texttt{results\_*\_routing.json}.}
The discipline: size the gate budget against a low percentile of the margin, and
protect the differential component; unlike keys, no DC term is needed --- relative
order is the geometry.

\textbf{Recurrences: the slow channels are fragile, and the error compounds.}
For a per-channel linear recurrence $h_t = a\,h_{t-1} + b_t$ the steady-state
gain is $1/(1-a)$ and the memory length is ${\sim}1/(1-a)$, so a fixed error in
the decay $a$ is amplified far more in slow (long-memory) channels ($a\to1$) and
accumulates over the sequence --- the recurrent analog of the RoPE-slow-channel
key finding, where the slow \emph{rotary} channels carried the fragile DC
offset. \texttt{decay\_sensitivity} returns the per-channel coefficient
$\sum_t t\,a^{t-1}$ (steady state $1/(1-a)^2$), which grows toward $a\to1$ and
with sequence length; on a synthetic recurrence a fixed decay error is amplified
$28$--$52\times$ more in slow than fast channels. The fix follows the geometry:
quantize the decay in a \emph{log-time-constant} basis ($\tau=-\log a$, then a
uniform grid on $\log\tau$), concentrating levels where $a\to1$ --- the SSM
analog of NF4's non-uniform key levels. \textbf{On real Mamba-790m}, quantizing
the continuous decay $A=-e^{A_{\log}}$ on a linear grid raises WikiText-2
perplexity from a baseline $11.65$ to $1.0\times10^{10}$ at 3-bit --- the slow
channels, whose $A$ ranges over many orders of magnitude, are annihilated ---
while quantizing in Mamba's native $A_{\log}$ (log) basis keeps it at $14.44$: a
gap of nearly nine orders of magnitude ($\sim\!7\times10^{8}$) from choosing the
basis the geometry dictates.\footnote{Measured by
\texttt{benchmarks/validate\_mamba\_decay.py} (fp32 eval so the collapsed variant
reports a finite perplexity rather than overflowing); raw data in
\texttt{results\_mamba\_decay.json}.} The synthetic $5$--$6\times$ becomes
catastrophic-versus-fine on a real model precisely because the real decay range
is enormous.

Both boundaries are measured on synthetic operators, and state decay now
end-to-end on a real SSM; the formal sensitivity theorems for gated selection
and selective recurrences are the theory paper's work, not claimed here.

\section{Certificates instead of promises}\label{sec:cert}
Track 1's quality story was, until recently, measured recall plus cosine ---
promises. The companion theory paper's rank proposition converts to a shipped
floor: for any corpus, measure the robust distortion $\kappa$ (97.5/2.5
percentile ratio of compressed to exact distance) and the corpus's
distance-ratio concentration $\hat\mu(\kappa)$ (one pass,
$O(P\log P)$); then \emph{with no distributional assumptions},
Kendall $\tau \geq 1-2\hat\mu(\kappa)$ and Spearman
$\rho \geq 1-3\hat\mu(\kappa)$ \cite{kta,daniels1950}. The floor's
\emph{vacuity} is itself the product feature: on distance-concentrated
corpora no finite distortion certifies rank --- exactly when single-stage
retrieval dies --- so ``exact reranking required'' becomes a derived,
per-corpus signal rather than a fixed oversampling menu; \texttt{autotune}
reports $\kappa$/$\hat\mu$/$\tau$-floor per operating point. Guarding the
other track: an (A2) \emph{consumer-metric probe} quantizes a calibration
batch under each candidate family's information quotient and compares rank
agreement of the declared consumer's scores (cosine, L2, attention logits)
--- it reproduces the \S\ref{sec:keys} catastrophe on synthetic key
statistics as a unit test --- and a streaming tangential-fraction statistic
in the quality monitor turns norm-dominated drift into a dashboard alert.
The probe work surfaced a distinction worth stating: there are \emph{two}
failure classes --- radial displacement (hubness; the streaming statistic
sees it) and direction concentration (the keys regime; displacements are
fully tangential yet squeezed below the quantizer's cell size --- only the
end-to-end probe sees it).

\section{Two empirical laws, with their boundaries}\label{sec:laws}
\textbf{Density dissociation.} A per-vector density quotient on
compressed-domain candidates (the retrieval analogue of the theory paper's
$\alpha{=}1$ normalization) helps \emph{iff} density is nuisance for the task
truth: $+15.7$ recall points --- beating exact reranking, which cannot remove
the nuisance --- when an anisotropic common mean is excluded from the ground
truth, and $-28$ points when it is part of it. The public-data replication
(BGE-M3 over WikiText-2, observed-space ground truth) confirmed both
predictions out-of-sample and isolated the shippable form: correct only the
\emph{compression-induced} density shift ($\mu_{\text{recon}} -
\mu_{\text{orig}}$, nuisance by construction, computable at index build):
$+0.3$ recall points where the plain quotient loses $5.9$.
\textbf{Spectral boundary for bit-tapering.} A pre-registered prediction from
the theory paper's heat-filter theorem --- exponentially tapered bit
schedules beat hard truncation at matched total bits --- held on a synthetic
power-law corpus (recall $+1.6$ pts at constrained budgets; certified
$\tau$-floor non-vacuous at half the budget) and \emph{reversed} on real
BGE-M3 embeddings. The reversal is quantified: the synthetic spectrum has
effective rank 13.6 (tail dims near-noise; cheap taper bits free), BGE has
136.7 (mid-spectrum dims carry real variance; coarse tail bits inject real
noise). The rule --- \emph{taper wins iff spectral effective rank is far
below the bit-budget's head width} --- is a one-number diagnostic from the
covariance the pipeline already estimates.

\section{Compute on the codes: fused decode}\label{sec:fused}
One idea serves both tracks: \emph{compute on the codes, rotate at the
boundaries}. In retrieval it is the ADC index. In decode it is a kernel
family that computes one attention step directly from compressed K/V ---
per-query LUT, online (flash) softmax over ADC scores, value accumulation in
code space, a single inverse rotation --- validated \emph{exact} against the
dequantize-then-attend path at every milestone and 1.8--13.3$\times$ faster
at 2k--32k context (split-K, one warp per (head, key-split), shuffle
reductions). New in this draft, the recommended \emph{key} format is fused
too: since keys enter attention only through $q\cdot k$, the whole
per-channel format becomes score terms ---
\[
\text{score}_s \;=\; \underbrace{q\cdot\mu(h)}_{\text{zero-point bias, folded once}}
\;+\; \underbrace{\textstyle\sum_j (q_j a_j)\,\mathrm{grid}[c_{js}]}_{\text{dense, branch-free}}
\;+\; \underbrace{\textstyle\sum_{j\in\mathrm{out}(s)} q_j\,\Delta_{sj}}_{\text{sparse CSR deltas}}
\]
with outliers stored token-major as \emph{deltas} against the dense dequant,
applied by a warp-cooperative pass (lanes stride the $\sim$3-entry row, one
shuffle reduction folds the correction in before the softmax update). Memory
contiguity comes from a one-off build-time channel-major$\to$token-major
transpose; branch divergence is bounded to the ragged tail of a short list
and never touches the dense loop. All three zero-point modes fold into the
same bias term --- the kernel never branches on the mode. Exact to
$8{\times}10^{-8}$ against decompress-then-attend across every key variant,
and 2--6$\times$ faster end-to-end even with per-call CSR rebuild (the cache
integration amortizes it per cold flush).

\section{Methodology as a contribution}\label{sec:method}
Three habits made the results above auditable. \textbf{Evidence ladder:}
every headline claim carries a level (L1 public one-click reproduction
through L5 unit-tested engineering) and links a runnable notebook; L4/L5 is
a disclosure, not a weakness. \textbf{Pre-registration:} directional
predictions (the taper test, the dissociation replication) are written down
before the run, so a reversal is a documented boundary rather than a quiet
omission. \textbf{Incident to instrument:} the keys catastrophe became the
consumer-metric probe; the GQA collapse became asym-NF4 and then the
deterministic zero-point; the cosine-vs-recall divergence became rank
certificates. The system's guarantees section exists because its failures
were kept.

\section{Related work}\label{sec:related}
TurboQuant (online vector quantization) \cite{turboquantpaper} supplies the
rotation-and-scalar-quantize primitive; QLoRA's NF4 \cite{qlora} the
codebook asym-NF4 repairs; KVQuant \cite{kvquant} the calibrated per-channel
NUQ that asym-NF4 matches calibration-free; RaBitQ/OPQ/PQ the retrieval
baselines under one harness. Varici et al.\ \cite{varici2025} justify the
PCA-Matryoshka rotation. Hubness correction in high-dimensional kNN
motivates \S\ref{sec:laws}'s quotient, with the dissociation supplying the
missing applicability condition. The companion theory paper \cite{kta}
provides the transfer theorems, the rank proposition, the (A2) condition,
and the heat-filter prediction tested here; the KV write-up \cite{kvtmlr}
details \S\ref{sec:nf4}--\ref{sec:rope}. Rababah et al.\ \cite{illusion}
introduce the equivalency framing and Correctness Agreement that
\S\ref{sec:behavioral} sharpens with a noise floor and whose
projection-sensitivity ranking it reverses at matched bits. Selective
state-space models (Mamba \cite{mamba}) and sparse mixture-of-experts routing
(Mixtral \cite{mixtral}) motivate \S\ref{sec:hybrid}; their quantization has been
studied empirically, but the operator-level sensitivity boundary --- the selection
margin for gates, the log-time-constant basis for decays --- is the (A2)
contribution here rather than a benchmarking sweep.

\section{Limitations and open problems}\label{sec:limits}
KV results cover four models $\leq$14B on LongBench-English/WikiText-2; the
bias zero-point is Qwen-family by construction; the fused per-channel kernel
awaits its cache-dispatch integration and the nuq grid stays
decompress-then-attend. The rank certificate is worst-case and typically far
below measured recall (a loose floor, honestly priced). The delta-centering
retrieval option and the effective-rank taper diagnostic are validated on
one public corpus each. The GQA-tolerance threshold rests on four
architecture points. The projection-sensitivity reversal (\S\ref{sec:behavioral})
is fake-quantization on three models over a next-token sample, directional
rather than absolute, and still owes the source paper's multiple-choice tasks
and the gated Llama-3.2-3B. The operator-regime tracer (\S\ref{sec:infer}) is
heuristic --- name/type rules plus a best-effort \texttt{torch.fx} pass, defeated
by dynamic control flow and models fx cannot symbolically trace, where it falls
back to the conservative structural default. The gate and state-decay boundaries
(\S\ref{sec:hybrid}) are confirmed end-to-end on real models --- routing on
Mixtral-8x7B-Instruct (top-2) and OLMoE-1B-7B (top-8), state decay on
Mamba-790m --- but on WikiText-2 only, and the routing evidence is margin-
concentration (whose absolute flip rate is gate-quantizer-dependent) rather than
a single calibrated deployment recipe; the broader multi-model sweep and the
formal sensitivity theorems for gated selection and selective recurrences are
still open. Deployment-grade format freezing (conformance
tests, fuzzing) is roadmapped but not complete.

\begin{thebibliography}{9}
\bibitem{kta} A.~H.~Bond. Keep the angle: a geometry-preserving basis in
spectral embeddings. Working draft v0.8,
\url{https://github.com/ahb-sjsu/the-angular-observer}, 2026.
\bibitem{obstheory} A.~H.~Bond. Observation Theory: geometry, distortion, and
reliability as properties of observation.
\url{https://doi.org/10.5281/zenodo.21423315}, 2026.
\bibitem{kvtmlr} A.~H.~Bond. One codebook for every architecture: asymmetric
NormalFloat for calibration-free KV-cache quantization. Draft,
\texttt{paper/kv\_tmlr/}, 2026.
\bibitem{turboquantpaper} A.~Zandieh et al. TurboQuant: online vector
quantization with near-optimal distortion. ICLR, 2026.
\bibitem{qlora} T.~Dettmers, A.~Pagnoni, A.~Holtzman, and L.~Zettlemoyer.
QLoRA: efficient finetuning of quantized LLMs. NeurIPS, 2023.
\bibitem{kvquant} C.~Hooper et al. KVQuant: towards 10 million context
length LLM inference with KV cache quantization. NeurIPS, 2024.
\bibitem{varici2025} B.~Varici et al. On the theory of Matryoshka
representations and PCA recovery of ordered eigenfunctions. 2025.
\bibitem{daniels1950} H.~E.~Daniels. Rank correlation and population models.
\emph{J.\ Royal Stat.\ Soc.\ B}, 12(2):171--181, 1950.
\bibitem{illusion} B.~Rababah, C.~G.~Akcora, and C.~K.~Leung. The illusion of
equivalency: statistical characterization of quantization effects in LLMs.
arXiv:2607.08734, 2026.
\bibitem{mamba} A.~Gu and T.~Dao. Mamba: linear-time sequence modeling with
selective state spaces. COLM, 2024.
\bibitem{mixtral} A.~Q.~Jiang et al. Mixtral of experts.
arXiv:2401.04088, 2024.
\end{thebibliography}

\end{document}
