\documentclass[11pt]{letter}
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}

\signature{Andrew H. Bond, Senior Member, IEEE\\
Department of Computer Engineering\\
San Jos\'{e} State University\\
andrew.bond@sjsu.edu}

\address{Andrew H. Bond\\
Department of Computer Engineering\\
San Jos\'{e} State University\\
San Jos\'{e}, CA 95192, USA}

\begin{document}

\begin{letter}{Editor-in-Chief\\
IEEE Transactions on Artificial Intelligence}

\opening{Dear Editor,}

I am submitting the manuscript ``PCA-Matryoshka: Enabling Effective
Dimension Reduction for Non-Matryoshka Embedding Models with
Applications to Vector Database Compression'' for consideration in
IEEE Transactions on Artificial Intelligence.

\textbf{Summary.}  Most deployed embedding models (BGE-M3, Cohere
Embed, older OpenAI models) cannot be effectively truncated for
compression because they were not trained with Matryoshka
representation learning.  We show that a training-free PCA rotation
recovers the truncation property, improving cosine similarity from
0.467 to 0.974 at 256 dimensions (109\% improvement).  Combined
with 3-bit scalar quantization, the pipeline achieves 27$\times$
compression at 0.979 cosine similarity on a 2.4-million-vector
corpus spanning 37 languages.

\textbf{Contributions.}
\begin{enumerate}
  \item A training-free technique (PCA-Matryoshka) that enables
  effective dimension truncation for any embedding model.
  \item The first systematic comparison of 15 embedding compression
  methods on a single large-scale corpus, identifying Pareto-optimal
  configurations.
  \item An open-source implementation deployed in production on
  3.3 million vectors.
\end{enumerate}

\textbf{Relevance to IEEE TAI.}  This work bridges the gap between
embedding model research and practical deployment constraints.  The
compression technique directly enables local AI deployment in
privacy-sensitive domains (healthcare, education, legal)---a growing
concern as AI regulation tightens globally.

The manuscript is fully anonymized for double-anonymous review.
It has not been submitted elsewhere and is not under consideration
at any other journal.

Thank you for your consideration.

\closing{Sincerely,}

\end{letter}
\end{document}
